### 2020-11-12 latex/acrobot.tex
Edited acrobot constraint and improved theorem

Defining \(M(q) := D_s(q)\) and \(V(q) := P_s(q)\), we find the conjugate of momenta 
is \(p = (p_u, p_a) = M(q)\dot{q}\).
The dynamics in \((q,p)\) coordinates are given by
\begin{align}\label{eqn:acrobot-hamiltonian}
    \mathcal{H}(q,p) &= \frac{1}{2}p\tpose \Minv(q) p -
    mgl\left(2 c_u + c_{ua}\right)
    , \\
     &\begin{cases}
        \dot{q} = \Minv(q) p \\
        \dot{p}_u = -mgl\left(2s_u + s_{ua}\right) \\
        \dot{p}_a =-\frac{1}{2}p\tpose \nabla_{q_a}\Minv(q) p
        - mgl s_{ua} + \tau,
    \end{cases} \nonumber
\end{align}
where the inverse inertia matrix is
\begin{equation}\label{eqn:Minv}
    \Minv(q) = \frac{1}{ml^2\left(2-c_a^2\right)}
    \begin{bmatrix}
        1 &
        -\left(1+c_a\right) \\
        -\left(1+c_a\right) &
        3+2c_a
    \end{bmatrix}
    .
\end{equation}
The control input is a force \(\tau \in \R\) affecting only the dynamics of
\(p_a\), representing a torque acting on the hip joint.
This means \((q,p)\) are simply actuated coordinates inside the phase space
\(\mathcal{Q} \times \mathcal{P}\) where
\[
    \mathcal{Q} = \mathcal{Q}_u \times \mathcal{Q}_a := 
    \mathbb{S}^1 \times \mathbb{S}^1
    ,
\]
and
\[
    \mathcal{P} = \mathcal{P}_u \times \mathcal{P}_a
    := \R \times \R
    .
\]
This allows us once again to apply the theory of VNHCs from Chapter
\ref{ch:vnhcs}.

Let us define the VNHC \(h(q,p) = q_a - f(q_u,p_u)\) of order 1, where
\(f \in C^2\left(\mathcal{Q}_u \times \mathcal{P}_u; \mathcal{Q}_a\right)\).
Since \(\nabla_{q_u}\Minv(q) = \Zmat{2\times 2}\),
Theorem \ref{thm:vnhc-regularity} tells us that this VNHC will be regular
if the regularity matrix
\[
    dh_q \Minv(q) \begin{bmatrix}0 \\ 1 \end{bmatrix}
    ,
\]
is of full rank on the constraint manifold \(\Gamma\).
Then, since 
\[
    dh_q = \begin{bmatrix}
        -\partial_{q_u} f & 1
    \end{bmatrix}
    ,
\]
the regularity matrix evaluates to the scalar equation
\begin{equation}\label{eqn:regularity-matrix-acrobot}
    \frac{(1+c_a)\partial_{q_u}f(q_u,p_u) + (3+2c_a)}{ml^2(2-c_a^2)}
    .
\end{equation}
This is full rank on \(\Gamma\) if and only if the numerator does not
change sign.

A sufficient condition for regularity is when \(f\) is a function solely of \(p_u\),
because then \(\partial_{q_u}f = 0\) and ~\eqref{eqn:regularity-matrix-acrobot}
is strictly positive for all values of \(q_a\). 
This will be useful later, as we will not need to check regularity if we design
a function of the unactuated momentum.

@@

One initial approach to controlling the acrobot is to model it as a
variable-length pendulum by collapsing the two rods and masses into one
equivalent center of mass (ECM), as in Figure \ref{fig:acrobot-ecm}.
This seems a reasonable model reduction, since the length from the pivot to the
ECM changes depending on the angle \(q_a\) of the leg.
Indeed, \citet{swingup_giant_acrobot} use this approach to design a trajectory
for the ECM, then determine which leg angles \(q_a(t)\) are required to generate
that trajectory.
Following in their footsteps, we might consider using the results
from Chapter \ref{ch:vlp} to find the leg angles that allow the ECM to gain
energy. 
Then we could apply Theorem \ref{thm:vlp-energy-stabilization} to prove
the acrobot is gaining energy too.
\begin{figure}
    \centering
    \includestandalone[width=0.5\textwidth]{images/acrobot_ecm}
    \caption{A simple acrobot modelled as a VLP with equivalent center of mass \(2m\). 
        The length of the VLP changes according to \(q_a\).}
    \label{fig:acrobot-ecm}
\end{figure}

@@

Second, if we were to compute the conjugate of momenta
\(p_{l_e}\) to \(l_e\) and \(p_e\) to \(q_e\), we would see the torque input
\(\tau\) appearing in both of their dynamic equations.
In the VLP model from Chapter \ref{ch:vlp}, the control input only
affects the dynamics of the length variable.
If we want to design a VNHC for this system, we cannot use any of the results
from Chapter \ref{ch:vlp} because the VLP models don't match.

@@

\section{The Acrobot Constraint}
% TODO: Give the constraint, state the theorem, and show the outline of the
% proof (both as a sketch and in a figure)
One may be tempted to design a constraint of the form \(q_a = \bar{q}_a\sin(\theta)\) 
with \(\theta := \arctan_2(p_u,q_u)\), since a similar approach 
was so effective for the VLP in Chapter \ref{ch:vlp}.
Unfortunately, this constraint is not regular, and it is difficult to find any
VNHC of the form \(q_a = f(\theta)\) where it is easy to prove regularity
analytically.
Instead, we will develop a constraint \(h(q,p) = q_a - f(p_u)\) which,
as we saw earlier, is always regular and is likely to inject energy.

@@

Now imagine the person's torso is affixed to the swing's rope so that they are
always upright. 
Imagine further that the swing has no seat at all, allowing the person to extend
their legs beneath them. 
This position is identical to that of a gymnast on a bar, which is why we can
use the leg motion from seated swinging to design a controller for the acrobot.

The acrobot's legs are rigid rods which cannot retract, so we emulate the person
on a swing by pivoting the legs in the direction of motion. 
To account for how a person leans back at higher speeds, the legs should pivot to an
angle proportional to the swing's speed.
One VNHC which emulates this process is \(q_a = \bar{q}_a\arctan_2( I p_u)\), as
in Figure \ref{fig:qa-arctan}.
Here, \(\bar{q}_a \in ]0,\frac{2 Q_a}{\pi}]\)
and \(I \in \R_{>0}\) is a fixed control parameter.

This constraint does not perfectly recreate a gymnast's motion -- during
a giant the gymnast's legs are almost completely extended, while this constraint
pivots the legs partially during rotations.
However, the motion looks similar enough that this constraint should provide a
decent foundation for injecting energy into the acrobot.
It is for this reason that we choose our acrobot's constraint to be
\begin{equation}\label{eqn:acrobot-constraint}
    h(q,p) = q_a - \bar{q}_a \arctan_2(I p_u)
    .
\end{equation}

\begin{figure}
    \centering
    \includestandalone[width=0.5\textwidth]{images/qa_arctan}
    \caption{The acrobot constraint \(q_a = \bar{q}_a \arctan_2(I p_u)\).}
    \label{fig:qa-arctan}
\end{figure}

@@

These dynamics do not always gain energy -- we need additional conditions
on the control parameters to guarantee this is true.

% TODO: This needs modification. The energy injection is provable for any system
% only during oscillations. For rotations, we need to do a numerical check of an
% integral that depends on m,g,l. If this integral is positive, then the system
% is gaining energy up to a point - it sems the integral decays to 0 so that you
% can reach a limiting speed, which means you don't get full energy injection
% everywhere on Gamma.
\begin{thm}\label{thm:acrobot-energy-stabilization}
    Consider the simple acrobot ~\eqref{eqn:acrobot-hamiltonian} constrained by
    ~\eqref{eqn:acrobot-constraint} with \(\bar{q}_a \in ]0,1]\).
    There exists a small enough\footnotemark \(I^\star \in \R_{>0}\)
    such that the VNHC injects energy whenever \( I \in ]0,I^\star[\). 
    If \(I \in ]-I^\star,0[\), the VNHC dissipates energy.
\end{thm}
\footnotetext{In fact, we can be more precise than this. 
    Letting \(\epsilon > 0\), we need 
    \(I \leq \min\left\{I_1(\epsilon),I_2\right\}\) where
    \(I_1(\epsilon) = \frac{\tan(\epsilon)}{(\pi-\epsilon)}
    \left(\frac{m^2gl^3(\cos(\epsilon)-2\sin(\epsilon))}{\pi-\epsilon}
    \left(\frac{2\bar{q}_a\tan(\epsilon)}{(1+\tan(\epsilon)^2)(\pi-\epsilon)}+5\right) 
    \right)^{-1/2}\)
    and \(I_2 = \sqrt{\frac{3}{10 m^2gl^3\bar{q}_a^2}}\).
    See Apendix \ref{apx:I-bounds} for details.
}

@@

There is a suitable change of coordinates into a pseudo-radius 
\(\mu \in \R_{>0}\) and pseudo-angle \(\alpha \in \mathbb{S}^1\) on the
\((q_u,p_u)\)-plane where \(\dot{\mu} = 0\) and \(\dot{\alpha} > 0\) when 
we set \(I = 0\).
Perturbation theory shows that, for small enough \(I\), the radius \(\mu\)
increases on average along \(\alpha\).
Hence, the acrobot is gaining energy since orbits in the \((q_u,p_u)\)-plane
spiral away from the origin, and the acrobot always begins performing rotations.

>>>>

Defining \(M(q) := D_s(q)\) and \(V(q) := P_s(q)\), we find the conjugate of momenta 
is \(p = (p_u, p_a) = M(q)\dot{q}\).
The dynamics in \((q,p)\) coordinates are given by
\begin{align}\label{eqn:acrobot-hamiltonian}
    \mathcal{H}(q,p) &= \frac{1}{2}p\tpose \Minv(q) p -
    mgl\left(2 c_u + c_{ua}\right)
    , \\
     &\begin{cases}
        \dot{q} = \Minv(q) p \\
        \dot{p}_u = -mgl\left(2s_u + s_{ua}\right) \\
        \dot{p}_a =-\frac{1}{2}p\tpose \nabla_{q_a}\Minv(q) p
        - mgl s_{ua} + \tau,
    \end{cases} \nonumber
\end{align}
where the inverse inertia matrix is
\begin{equation}\label{eqn:Minv}
    \Minv(q) = \frac{1}{ml^2\left(2-c_a^2\right)}
    \begin{bmatrix}
        1 &
        -\left(1+c_a\right) \\
        -\left(1+c_a\right) &
        3+2c_a
    \end{bmatrix}
    .
\end{equation}
The control input is a force \(\tau \in \R\) affecting only the dynamics of
\(p_a\), representing a torque acting on the hip joint.
This means \((q,p)\) are simply actuated coordinates inside the phase space
\(\mathcal{Q} \times \mathcal{P}\) where
\(\mathcal{Q} = \mathcal{Q}_u \times \mathcal{Q}_a 
:= \mathbb{S}^1 \times \mathbb{S}^1\), and
\(\mathcal{P} = \mathcal{P}_u \times \mathcal{P}_a
:= \R \times \R\).
This allows us to apply the theory of VNHCs from Chapter
\ref{ch:vnhcs}.

Let us define the VNHC \(h(q,p) = q_a - f(q_u,p_u)\) of order 1, where
\(f \in C^2\left(\mathcal{Q}_u \times \mathcal{P}_u; \mathcal{Q}_a\right)\).
Since \(\nabla_{q_u}\Minv(q) = \Zmat{2\times 2}\),
Theorem \ref{thm:vnhc-regularity} tells us that this VNHC will be regular
when the regularity matrix
\[
    dh_q \Minv(q) \begin{bmatrix}0 \\ 1 \end{bmatrix}
    ,
\]
is of full rank 1 on the constraint manifold \(\Gamma\).
Given that
\(dh_q = \begin{bmatrix} -\partial_{q_u} f & 1 \end{bmatrix}\),
the regularity matrix evaluates to the scalar equation
\begin{equation}\label{eqn:regularity-matrix-acrobot}
    \frac{(1+c_a)\partial_{q_u}f(q_u,p_u) + (3+2c_a)}{ml^2(2-c_a^2)}
    .
\end{equation}
This is full rank on \(\Gamma\) if and only if the numerator does not
change sign.
The following proposition provides a sufficient condition for this to happen.

\begin{prop}\label{prop:acrobot-fpu-regular}
    A relation \(h(q,p) = q_a - f(p_u) = 0\) for ~\eqref{eqn:acrobot-hamiltonian} 
    with \(f \in C^2\left(\mathcal{P}_u; \mathcal{Q}_a\right)\) is a regular
    VNHC of order 1.
\end{prop}
\begin{proof}
    Since \(\partial_{q_u} f = 0\), the regularity equation
    ~\eqref{eqn:regularity-matrix-acrobot} is strictly positive for all values
    of \(q_a\), and hence is of full rank 1.
    By Theorem \ref{thm:vnhc-regularity}, \(h\) is a regular VNHC of order 1.
\end{proof}
Proposition \ref{prop:acrobot-fpu-regular} will be useful later, as we will not
need to check regularity if we design a function of the unactuated momentum.

@@

One initial approach to controlling the acrobot is to model it as a
variable-length pendulum by collapsing the two rods and masses into one
equivalent center of mass (ECM), as in Figure \ref{fig:acrobot-ecm}.
This seems a reasonable model reduction, since the length from the pivot to the
ECM changes depending on the angle \(q_a\) of the leg.
Indeed, \citet{swingup_giant_acrobot} use this approach to design a trajectory
for the ECM, then determine which leg angles \(q_a(t)\) are required to generate
that trajectory.
Following in their footsteps, we might consider using the results
from Chapter \ref{ch:vlp} to find the leg angles that allow the ECM to gain
energy. 
Then we could apply Theorem \ref{thm:vlp-energy-stabilization} to prove
the acrobot is gaining energy.
\begin{figure}
    \centering
    \includestandalone[width=0.5\textwidth]{images/acrobot_ecm}
    \caption{A simple acrobot modelled as a VLP with equivalent center of mass \(2m\). 
        The length of the VLP changes according to \(q_a\).}
    \label{fig:acrobot-ecm}
\end{figure}

@@

Second, if we were to compute the conjugate of momenta
\(p_{l_e}\) to \(l_e\) and \(p_e\) to \(q_e\), we would see the torque input
\(\tau\) appearing in both of their dynamic equations.
In the VLP model from Chapter \ref{ch:vlp}, the control input only
affects the dynamics of the length variable.
If we want to design a VNHC for this system, we cannot use any of the results
from Chapter \ref{ch:vlp} because the VLP models do not match.

@@

\section{The Acrobot Constraint}
% TODO: Give the constraint, state the theorem, and show the outline of the
% proof (both as a sketch and in a figure)
One may be tempted to design a constraint of the form \(q_a = \bar{q}_a\sin(\theta)\) 
with \(\theta := \arctan_2(p_u,q_u)\), since a similar approach 
was so effective for the VLP in Chapter \ref{ch:vlp}.
Unfortunately, this constraint is not regular, and it is difficult to find any
VNHC of the form \(q_a = f(\theta)\) where regularity can be proven easily.
Instead, we will develop a constraint \(h(q,p) = q_a - f(p_u)\) because these
constraints are always regular (as per Proposition \ref{prop:acrobot-fpu-regular}).

@@

Now imagine the person's torso is affixed to the swing's rope so that they are
always upright. 
Imagine further that the swing has no seat at all, allowing the person to extend
their legs beneath them. 
This position is identical to that of a gymnast on a bar, which is why we can
use leg motion from the seated swing to design a controller for the acrobot.

The acrobot's legs are rigid rods which cannot retract, so we emulate the person
on a swing by pivoting the legs in the direction of motion. 
To account for how a person leans back at higher speeds, the legs should pivot to an
angle proportional to the swing's speed.
One VNHC which emulates this process is \(q_a = \bar{q}_a\arctan( I p_u)\), as
in Figure \ref{fig:qa-arctan}.
Here, \(\bar{q}_a \in ]0,\frac{2 Q_a}{\pi}]\)
and \(I \in \R_{>0}\) is a fixed control parameter.

This constraint does not perfectly recreate a gymnast's motion -- during
a giant the gymnast's legs are almost completely extended, while this constraint
pivots the legs partially during rotations.
However, the motion looks similar enough that this constraint should provide a
decent foundation for injecting energy into the acrobot.
It is for this reason that we choose our acrobot's constraint to be
\begin{equation}\label{eqn:acrobot-constraint}
    h(q,p) = q_a - \bar{q}_a \arctan(I p_u)
    .
\end{equation}

\begin{figure}
    \centering
    \includestandalone[width=0.5\textwidth]{images/qa_arctan}
    \caption{The acrobot constraint \(q_a = \bar{q}_a \arctan(I p_u)\).}
    \label{fig:qa-arctan}
\end{figure}

@@

These dynamics do not always gain energy -- we need \(|I|\) to be small enough
to guarantee this is true.

\begin{thm}\label{thm:acrobot-energy-stabilization}
    Consider the simple acrobot ~\eqref{eqn:acrobot-hamiltonian} constrained by
    the VNHC ~\eqref{eqn:acrobot-constraint}, whose
    constraint manifold is \(\Gamma = \mathbb{S}^1 \times \R\).
    Let \(E\) be the nominal energy when \(I = 0\):
    \[
        E(q_u,p_u) := \frac{p_u^2}{10ml^2} + 3(1 - \cos(q_u))
        .
    \]
\begin{enumerate}
    \item There always exists \(I \in \R_{>0}\) small enough that 
    ~\eqref{eqn:acrobot-constraint} injects energy into the acrobot on the set
    \[
        \mathcal{O} := \left\{(q_u,p_u) \in \Gamma 
            \mid E(q_u,p_u) < E(\pi,0) \right\}
        .
    \]
    If instead \(I \in \R_{<0}\), the VNHC dissipates energy on \(\mathcal{O}\).
\item Define \(b : \mathbb{S}^1 \times \R_{> 0} \rightarrow \R\) by
    \[
        b(\alpha,\mu_0) := 
        \frac{5m^2 g l^3 \bar{q}_a \left(
            m^2gl^3\left(18s_\alpha^2 + 30c_\alpha(1 - c_\alpha)\right)
            - c_\alpha\mu_0^2
        \right)}{
        |\mu_0|\sqrt{\mu_0^2 - 30m^2gl^3(1 - c_\alpha)}
        }
        ,
    \]
    and fix \(\bar{\mu} > 60m^2gl^3\).
    Suppose \(\int \limits_{0}^{2\pi} b(\sigma,\mu_0)d\sigma > 0\) for all 
    \(\mu_0 \in ]60m^2gl^3, \bar{\mu}[\).
    Then there exists \(I \in \R_{> 0}\) small enough that
    ~\eqref{eqn:acrobot-constraint} injects energy into the acrobot on
    \[
        \Omega := \left\{(q_u,p_u) \in \Gamma 
            \mid E(q_u,p_u) < E(0,\bar{\mu})\right\}
        .
    \]
    If instead \(I \in \R_{<0}\), the VNHC dissipates energy on \(\Omega\).
\end{enumerate}
\end{thm}
%\footnotetext{In fact, we can be more precise than this. 
%    Letting \(\epsilon > 0\), we need 
%    \(I \leq \min\left\{I_1(\epsilon),I_2\right\}\) where
%    \(I_1(\epsilon) = \frac{\tan(\epsilon)}{(\pi-\epsilon)}
%    \left(\frac{m^2gl^3(\cos(\epsilon)-2\sin(\epsilon))}{\pi-\epsilon}
%    \left(\frac{2\bar{q}_a\tan(\epsilon)}{(1+\tan(\epsilon)^2)(\pi-\epsilon)}+5\right) 
%    \right)^{-1/2}\)
%    and \(I_2 = \sqrt{\frac{3}{10 m^2gl^3\bar{q}_a^2}}\).
%    See Apendix \ref{apx:I-bounds} for details.
%}
The region \(\mathcal{O}\) is the oscillation domain of the simple pendulum
obtained by setting \(q_a = 0\);
\(\Omega\) includes \(\mathcal{O}\), alongside all rotations of this 
pendulum with momentum less than \(\bar{\mu}\). 
Hence, the first result of Theorem \ref{thm:acrobot-energy-stabilization} states
that the acrobot will gain enough energy to begin performing giant motion;
the second result guarantees can achieve giants with momentum
\(|p_u| < \bar{\mu}\).

@@

There is a suitable change of coordinates into a pseudo-radius 
\(\mu \in \R_{>0}\) and pseudo-angle \(\alpha \in \mathbb{S}^1\) on the
\((q_u,p_u)\)-plane.
These coordinates have the property that \(\dot{\alpha} > 0\) when
\(I\) is small enough.
Perturbation theory shows that (for possibly smaller \(I\)) the radius \(\mu\)
increases on average along \(\alpha\).
Hence, orbits in the \((q_u,p_u)\)-plane spiral away from the origin, 
which means the acrobot is gaining energy.
